---
title: "It's about time"
subtitle: "Fairness and temporal depth"
author: "Joshua Loftus (LSE Statistics)"
output:
  xaringan::moon_reader:
    css: ["default", "xaringan-themer.css",  "ninjutsu", "extra.css"]
    seal: true
    lib_dir: libs
    nature:
      titleSlideClass: ["bottom", "left"]
      countdown: 59000
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
header-includes: 
  - \usepackage{tikz}
  - \usepackage{color}      
---

class: split-four

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.retina = 3, warning = FALSE, message = FALSE)
options(knitr.table.format = "html")
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#2d708e",
  secondary_color = "#230433",
  link_color = "#55c667",
  text_bold_color = '#f68f46',
#  title_slide_text_color = "#444444",
  title_slide_background_color = "#ffffff", #"#042333",
  title_slide_background_image = "../../../files/theme/LSE/LSEcrest_large.jpg",
  title_slide_background_size = "cover",
  ) #or contain
```

```{r xaringanextra, include=FALSE, warning=FALSE}
#library(xaringanExtra)
#xaringanExtra::use_animate_all("slide_left")
#xaringanExtra::use_tile_view()
#xaringanExtra::use_panelset()
#xaringanExtra::style_panelset_tabs(font_family = "inherit")
```

```{r tidyverse, include=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
theme_set(theme_minimal(base_size = 22))
set.seed(1)
library(broom)
library(modelr)
```

<style type="text/css">
.remark-slide-content {
    font-size: 1.2rem;
    padding: 1em 4em 1em 4em;
}
</style>

#### Problem

- Lots of fairness definitions (at least 21!)
- Impossibility/trade-offs between them
- What do?
- How to think about this choice?

--

#### Time heuristic
"Center" an oft-ignored dimension: **time**

(i.e. consider the times when variable values are determined)

---

#### Notation for fair classification

Desired outcome $Y = 1$, sensitive attribute $A$, predictor $M$ (mediator/merit), algorithm score/classification $D$

#### Causal graph: time flows left to right
.left-column[
Note: arrows from $A$ may be due to discrimination (e.g. $M$ could be an unfair measure of "merit")
]
.right-column[
<center>
```{r echo = FALSE}
library(DiagrammeR)
mermaid("
graph LR
  subgraph  
  A((A))-->M((M))
  M-->Y((Y))
  end
  Y-.->D
  A-->Y
  A-.->D
  M-.->D
")
```

</center>
]


---

#### Demographic parity 
Predictions (or decisions) are independent of $A$:
$$
\mathbb P(D = 1 | A = 0) = \mathbb P(D = 1 | A = 1)
$$


#### Equality of opportunity (Hardt et al, 2016)
Conditional parity, among individuals 'qualified' for the desirable outcome
$$
\mathbb P(D = 1 | A = 0, Y = 1) = \mathbb P(D = 1 | A = 1, Y = 1)
$$

#### Causal versions (e.g. counterfactual fairness)
Pathway analysis, *allowing* a causal pathway to change the prediction is like conditioning on a variable in that pathway

---

### Allowed variables

Definition: Variable $W$ determined at time $t$ is "allowed" by fairness definition $C$ if:
  - $C$ is a conditional fairness definition and $W$ is in the condition
  - $C$ is a causal pathway definition and $W$ is on an unblocked (i.e. resolving) pathway

---

### Temporal depth heuristic


#### Weak heuristic
If $W$, determined at time $t$, is allowed by $C$ then $C$ "justifies" any unfairness associated with (or causally flowing through) $W$ occurring before $t$

#### Strong heuristic (thick/dense causality)
If $W$, determined at time $t$, is allowed by $C$ then $C$ "justifies" all unfairness occurring before time $t$


---

### Applying the time heuristic

According to the strong heuristic, *equality of opportunity essentially allows all unfairness*

Demographic parity or (non-path-specific) counterfactual fairness do not justify any unfairness

(Future work tbd)

---

### Why allow any unfairness?

EU [Equality Directive](https://eur-lex.europa.eu/legal-content/EN/ALL/?uri=CELEX:32000L0043), CHAPTER I, Article 2 (b):

> indirect discrimination [...] would put persons of a racial or ethnic origin at a particular disadvantage compared with other persons, *unless that provision, criterion or practice is objectively **justified*** by a legitimate aim and the means of achieving that aim are appropriate and necessary.

US [civil rights](https://www.justice.gov/crt/fcs/T6Manual7#P) law: 

> If the evidence establishes a prima facie case of adverse disparate impact [...] courts then determine whether the recipient has articulated a "*substantial legitimate **justification***" [...]

---

### Maybe the law is bad?

Disclaimer: I am not a lawyer

**The end** 

(please talk to me)