---
title: "Discussion of Panigrahi et al"
subtitle: ""
author: "Joshua Loftus (LSE Statistics)"
#institute: "LSE"
#date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    css: ["default", "xaringan-themer.css", "../../../theme.css", "ninjutsu"]
#    seal: false    
    lib_dir: libs
    nature:
      titleSlideClass: ["bottom", "left"]
      countdown: 59000
      highlightStyle: github
      highlightLines: true
      highlightSpans: true
      countIncrementalSlides: false
header-includes: 
  - \usepackage{tikz}
  - \usepackage{color}
---


```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(fig.retina = 3, warning = FALSE, message = FALSE)
options(knitr.table.format = "html")
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_duo_accent(
  primary_color = "#2d708e",
  secondary_color = "#230433",
  link_color = "#55c667",
  text_bold_color = '#f68f46',
#  title_slide_text_color = "#444444",
  title_slide_background_color = "#ffffff", #"#042333",
  title_slide_background_image = "../../files/theme/LSE/sign_reflection.jpg",
  title_slide_background_size = "cover",
  ) #or contain
```

```{r xaringanextra, include=FALSE, warning=FALSE}
library(xaringanExtra)
#xaringanExtra::use_animate_all("slide_left")
xaringanExtra::use_tile_view()
xaringanExtra::use_panelset()
xaringanExtra::style_panelset_tabs(font_family = "inherit")
```

```{r tidyverse, include=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
theme_set(theme_minimal(base_size = 22))
set.seed(1)
library(broom)
library(modelr)
```

<style type="text/css">
.remark-slide-content {
    font-size: 1.2rem;
    padding: 1em 4em 1em 4em;
}
</style>

### Conflict of interest

--

I really like this work!

- Compelling applications of geometry 

Proof (for skeptical students?) it's worth "leveling up" on multivariate calculus, etc

Would love to see more diagrams (even heuristic ones)

- Interesting bridge/blending of frequentist and Bayesian motivations and methods

Philosophers of science are gonna have a great time trying to figure out what we're doing here (job creation?)

---

#### An extreme(?) thought experiment

- Population of scientists analyzing their datasets

- With probability $1-p$, undetected formatting error $\to$ data read incorrectly $\to$ selected model is nonsense

Would each scientist want to know if this has occurred in their analysis?

Obviously! BUT...

--

Mathematical assumptions of the inference method may still (approximately) hold (depending on the kind of reading error) $\to$ inference may still be valid

What's the problem? **Wrong parameters/hypotheses**

---

#### A less extreme(?) thought experiment

- With probability $1-p$, the group lasso selects a model which is **importantly wrong** (Box's term)

- Maybe the inference is still valid, but for parameters that are not of true scientific interest

Would each scientist want to know if this has occurred in their analysis?

Obviously! BUT...

--

It's essentially impossible for any method to guarantee. Model selection is hard!

Model diagnostics, stability selection, **adjustment factor**?

---

### Content warning: null hypothesis significance testing

Some in the audience might be thinking, like me: "*There's a hypothesis test on the other side of them there intervals*"

- What is the null?

- What is the (powered) alternative?

--

Are these harder to interpret for *overlapping group lasso* or other special cases?

---

### Content warning: informative(?) priors

Many wise statisticians would not object to (informative) priors, of course, understanding balance of dependence on *both* assumptions and data

- Paper shows concentration of "the" posterior if $n \to \infty$ and $p$ is **fixed**

--

But we often rely on model selection in high-dimensions with $p > n$

Does the proposed methodology "work" in that case?

---

### Question recap

**Selection probability**: can the adjustment factor be used for inference *about the selection*?

**Null hypothesis**: is there anything meaningful about an interval containing zero?

**Prior dependence**: how strong is the dependence on priors?

**Cost of validity/statistical efficiency**: computational efficiency?


